 

### 2021-02-02 vertical-thermal-review.Rmd
Minor edits and comments starting at line 260

Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light. However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area- based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al. 2008]. In a temperate Acer saccharum forest, within-canopy Acer saccharum exhibited greater LMA and $N area$ with height early in the growing season as a result of hydrostatic constraints on leaf morphology; however, in later growing season LMA and $N area$ continued to increase in the upper canopy due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016].In addition, carbon isotope composition also increased with height, independent of light [@cobleLightAcclimationOptimizes2015]

Height-based leaf trait variation in *Quercus alba* and *Quercus velutina* showed decrease in lobation with increased height in two tall species, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020]. However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus spp.

@@

Several studies point towards common variations and differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

@@

Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al. 2019; Tserej & Feeley 2021]. Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017]. With increase in effective leaf width, boundary layer thickness and resistance increases along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018]. Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al. 2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$. Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref]. Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].

@@

High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition through enabling D1 protein replacement. Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis--photoprotective pigments comprising of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. These are found to be proportional to within-canopy light gradient at an inter- and intra-species level. Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998]. In neotropical forest, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pool on an area or chl basis compared to low light leaves that showed greater mass based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al. 2009]

@@

##### lifespan
##### Deciduousness

@@

The combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees. For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc.. whereas understory leaves are subjected to shade or limitation of light, niche competition and herbivory. Overstory sun leaves are subjected to overheating pressures; thus, through developed traits of increased stomatal density, transpiration and water supply, along with increased convective heatloss via reduced boundary layer resistance, sun leaves are able to maintain optimum temperatures for photosynthesis and metabolic processes. Additionally, overstory sun leaves tend to have long-lived leaves[*double check*] with high LMA while shade leaves of canopy trees have large leaves with low LMA. Adaptive traits in understory enable light capture efficiency while maintaining larger width leaves with comparatively lower stomatal density due to low levels of photosynthesis in the shade, lesser $VPD$, and wind speed. Those differences in traits are likely to affect their thermal properties as well. Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [@rozendaalPlasticityLeafTraits2006; @niinemetsResponsesForestTrees2010a]

@@

**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`).
Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020] 
$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [@michaletzPlantThermoregulationEnergetics2015].

@@

Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm). Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance. However, counteracting this, lower wind speed in understory would reduce latent heat loss. Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance. However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018]. Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.

@@

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
The vertical gradient in photosythetic rates is also influenced by available concentration of N and P. (and water availability)
.[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]

@@

## Implications

*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*

@@

Models representing ecosystem dynamics related to carbon, energy, and water fluxes operate at a range of scales and have varying degrees of complexity , and include offline models and models that can be fully coupled with Earth System Models (ESMs). Dynamic global vegetation models (DGVMs) comprise the land surface models in ESMs used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks (Cao & Woodward, 1998; Foley et al., 1996; Sitch et al., 2003; Woodward & Lomas, 2004). Improved representation of vegetation demographic processes in ESMs, specifically including forest canopy architectural variation in both horizontal and vertical directions, has repeatedly been identified as a critical step toward reduced uncertainty  and more accurate characterization of biologically mediated feedbacks (Banerjee & Linn 2018, Evans, 2012; Moorcroft, 2006; Moorcroft, Hurtt, & Pacala, 2001; Purves & Pacala, 2008; Thomas, Brookshire, & Gerber, 2015). Still, as models increase in complexity, model developments and improvements have direct tradeoffs with increased computational cost and potentially decreased interpretability of model output.

>>>>

Light based acclimation in leaf traits is well established, while research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from changes in light availability. However, studies that have measured LMA in relation to height have found that in tropical forests LMA-height predicted stronger correlation between area-based respiration versus foliar nutrients, whereas Q10 and E10 remained constant with height [Cavaleri et al. 2008]. In a temperate Acer saccharum forest, ~~within-canopy Acer saccharum exhibited greater~~LMA and $N_{area}$ were greater with height early in the growing season as a result of hydrostatic constraints on leaf morphology. Later in the growing season, LMA and $N_{area}$ continued to increase in the upper canopy, but due to light driven changes [@cobleLightDrivesVertical2014a; @cobleLightAcclimationOptimizes2015; @cobleHowVerticalPatterns2016]. In addition, carbon isotope composition has also been shown to increase with height, independent of light [@cobleLightAcclimationOptimizes2015]

Height-based leaf trait variation in two tall temperate species, *Quercus alba* and *Quercus velutina*, showed decreases in lobation with increased height, suggesting an adaptation to hydraulic constraints in order to lower evapotranspiration with increasing height by decreasing lobation [@kusiPlasticLeafMorphology2020]. However, @sackHowStrongIntracanopy2006 did not observe trait variation based on height in Quercus species.

@@

Several studies point towards ~~common variations and~~differences in leaf trait plasticity within-species and at an intra-canopy gradient which influences leaf structural and biochemical traits [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. The differences among these traits contribute to leaf heat dissipation strategies and leaf temperature across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

@@

Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al. 2019; Tserej & Feeley 2021]. Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperature ranges [$T_{range}$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017]. With increases in effective leaf width, boundary layer thickness and resistance increase along with stomatal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018]. This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018]. Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissipate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Using leaf energy balance models to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al. 2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$. Leaf width and characteristic dimension are generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref]. Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation. Leaf lobes are deeper and leaf elongation and serration are more pronounced in sun exposed canopies than in the shaded understory~~, as well as leaf elongation and serration are more pronounced in sun exposed canopies~~ [@vogelConvectiveCoolingLow1970; @schueppTansleyReviewNo1993; @sackHowStrongIntracanopy2006].

@@

High light stimulates the need for photoprotection in leaves in order to protect internal photosystem II from chronic photoinhibition through enabling D1 protein replacement. Thus, leaves exposed to greater irradiance have greater pools of xanthophyll pigments expressed on a mass or area basis. Xanthophylls are photoprotective pigments comprised of violaxanthin, antheraxanthin and zeaxanthin [$VAZ$] that convert excess light into heat energy dissipation through non photochemical quenching of excitation energy, enabling short-term acclimation to high temperature and light [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. These are found to be proportional to within-canopy light gradients at an inter- and intra-species level. Therefore, $VAZ$ is lower in shaded leaves and greater in sun leaves, as observed among temperate tree species [@niinemetsAcclimationHighIrradiance1998]. In neotropical forests, foliage exposed to higher irradiance in the canopy showed greater b-carotene and vioxanthin-cycle pools on an area or chlorophyll basis compared to low light leaves that showed greater mass-based chlorophyll, lutein and neoxanthin carotenoids [Mastubara et al. 2009]

@@

##### Leaf lifespan [Elsa: phenology more broadly? - leaf lifespan, fruiting, flowering links to temperature and thermal gradients?]
##### Deciduousness

@@

The combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees. For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc.. whereas understory leaves are subjected to shade or limitation of light, niche competition and herbivory. Overstory sun leaves are subjected to overheating pressures; thus, through developed traits of increased stomatal density, transpiration and water supply, along with increased convective heat loss via reduced boundary layer resistance, sun leaves are able to maintain optimum temperatures for photosynthesis and metabolic processes. Additionally, overstory sun leaves tend to have long-lived leaves[*double check*] with high LMA while shade leaves of canopy trees have large leaves with low LMA. Adaptive traits in understory enable light capture efficiency while maintaining larger width leaves with comparatively lower stomatal density due to low levels of photosynthesis in the shade, lesser $VPD$, and wind speed. Those differences in traits are likely to affect their thermal properties as well. Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [@rozendaalPlasticityLeafTraits2006; @niinemetsResponsesForestTrees2010a]

@@

**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`).
Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [@michaletzPlantThermoregulationEnergetics2015].

@@

Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm). Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance. However, counteracting this, lower wind speed in understory would reduce latent heat loss. Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance. However, increases in air temperature typically also increase VPD. Resulting decreases in stomatal conductance, typically during midday, lead to a decrease in leaf CO2, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018]. Thus, $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.

@@

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation levels differ between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
The vertical gradient in photosythetic rates is also influenced by available concentration of N and P. (and water availability)
.[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]

@@

## Implications [ & future research directions?]

*Elsa: If emphsaize future research directions, could move the remote sensing opportunities to this section and frame modeling section around opportunities to use models to further explore open questions, scale these findings in space and time, and influence predictions*
*(In this section, we consider implications of the patterns reviewed above for our understanding of the future of forests--both how we model them and what sorts of responses we can expect under climate change.)*

@@

Models representing ecosystem dynamics related to carbon, energy, and water fluxes operate at a range of scales and have varying degrees of complexity, including offline models and models that can be fully coupled with Earth System Models (ESMs). Dynamic global vegetation models (DGVMs) comprise the land surface models in ESMs used to predict the global distribution of vegetation types and biosphere-atmosphere feedbacks (Cao & Woodward, 1998; Foley et al., 1996; Sitch et al., 2003; Woodward & Lomas, 2004). Improved representation of vegetation demographic processes in ESMs, specifically including forest canopy architectural variation in both horizontal and vertical directions, has repeatedly been identified as a critical step toward reduced uncertainty  and more accurate characterization of biologically mediated feedbacks (Banerjee & Linn 2018, Evans, 2012; Moorcroft, 2006; Moorcroft, Hurtt, & Pacala, 2001; Purves & Pacala, 2008; Thomas, Brookshire, & Gerber, 2015). Still, as models increase in complexity, model developments and improvements have direct tradeoffs with increased computational cost and potentially decreased interpretability of model output.
